---
title: "A brief guide to CAFT"
author: "Glen A Satten, Mo Li, Ni Zhao"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    toc: true
vignette: >
  %\VignetteIndexEntry{A brief guide to changepointGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The **CAFT** package develop a rank-based compositional analysis using the accelerate failure time model for Micorbiome data analysis.

```{r}
library(CAFT)
```


## Colon data.

### Extrac data set 

The data set is included in the R package "curatedMetagenomicData" and the package installation is required through "Bioconductor".

```{r}
data(Colon)
```

### Fitering OTU data table

```{r}
library(phyloseq)

count.tab = t(as.data.frame(as.matrix(otu_table(Colon))))
sample.tab = as.data.frame(as.matrix(sample_data(Colon)))
tax.tab = as.data.frame(as.matrix(tax_table(Colon)))

p = sample.tab$study_name %in% "WirbelJ_2018"
sample.tab = sample.tab[p,]
count.tab = count.tab[p, ]

dim(sample.tab)
dim(count.tab)

pNA = which(is.na(sample.tab$age))
if(length(pNA) > 0){
  count.tab = count.tab[-pNA, ]
  sample.tab = sample.tab[-pNA,]
}

# No missing values from gender

### otu presence filtering
p_otu = which(rowSums(t(count.tab) > 0) > 1)
count.tab = count.tab[,p_otu]
tax.tab = tax.tab[p_otu,]

```


### The of-interested and adjusting covariates

```{r}
Disease = as.numeric(factor(sample.tab$disease, 
                            levels = c("healthy", "CRC"))) - 1
Age = as.numeric(sample.tab$age)
Gender = as.numeric(factor(sample.tab$gender)) - 1
```

### Censoring proportion

```{r}
##### 1. OTU table dimension (86 samples and 369 identified taxa)
dim(count.tab)

##### 2. mean censoring proportion
cens.prop = colMeans(count.tab == 0, na.rm = T)
mean(cens.prop)
```

### Three AFT model based individual taxa hypothesis testing method

```{r}
res.CAFT = caft(otu.table=count.tab, Y=Disease, 
               C=data.frame(Age=Age, Gender=Gender),
               filter.thresh=0.06, CAFT.method="CAFT",
               adjust.method="BH")
```

Note that we can also calculate the censoring proportion (proportion of zero 
cells) for the filtered data.

```{r}
mean(cens.prop[res.AFT$skip.otu == 0])
```

## References

Pasolli, E., Schiffer, L., Manghi, P., Renson, A., Obenchain, V., Truong, D., Beghini, F., Malik, F., Ramos, M., Dowd, J., Huttenhower, C., Morgan, M., Segata, N., & Waldron, L. (2017). Accessible, curated metagenomic data through ExperimentHub. *Nature Methods*, 14(11), 1023–1024.


```{r}
sessionInfo()
```
